---
title: "DS 2870: Module 2 Homework - Used Cars"
author: "Joey Gilmartin"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE, message = F, warning = F}
knitr::opts_chunk$set(echo = TRUE,
                      #warning = F,
                      #message = F,
                      fig.align = "center")

# Load the needed package(s) here
pacman::p_load(tidyverse)
```

## Question 1: Box Plot for Olympian Ages

The data set **olympics.csv** (found at https://raw.githubusercontent.com/Shammalamala/DS-2870-Data-Sets/main/olympics.csv) has data on about 6000 Olympic athletes that completed in 2024 Olympic games in one of 10 sports:

Athletics, Swimming, Rowing, Judo, Shooting, Sailing, Volleyball, Equestrian, Fencing, Boxing, Cycling Road, Gymnastics

(Athletics is a catchall for Track and Field style of events)

The two relevant columns are:

1) *sport*: Which of the 10 sports the athlete participated in

2) *age*: The age the athlete is at the start of the 2024 Olympic games

Using the data set, create the side-by-side box plots seen in Brightspace. The hex codes for the colors are #0081c8 and #FCB131.

To reorder the sports to match what is in Brightspace, use `fct_reorder()` (to see how it works, the help menu is your friend!)


```{r Q1}
olympics <- read_csv('https://raw.githubusercontent.com/Shammalamala/DS-2870-Data-Sets/main/olympics.csv')
#make plot
gg_olympics <- ggplot(
  data = olympics,
  mapping = aes(
    x = age,
    y = fct_reorder(sport, olympics$age),
  )
) + #add boxplots and colors
  geom_boxplot(
    fill = "#0081c8",
    color = "#FCB131"
  ) + 
  theme_bw() + #base theme
  theme(
    plot.title = element_text(hjust = .5, face = 'bold'), #theme updates, bold title and center title and subtitle
    plot.subtitle = element_text(hjust = .5)
    
  ) +
  labs( 
    x = NULL, #update axis, title, and subtitles
    y = "Sport",
    title = "Age of Olympians at the start of the Olympics by Sport",
    subtitle = "10 of the most common sports only"
  )

gg_olympics

```

## Question 2: Used Cars

The **used cars.csv** file has info about 400 cars listed on Craigslist in 2023 (github link: https://raw.githubusercontent.com/Shammalamala/DS-2870-Data-Sets/main/used%20cars.csv)

The columns are: 

1) *id*: The unique ID for the car in the data set
2) *price*: The asking price of the used car on Craigslist (in dollars)
3) *year*: The year the car was manufactured
4) *manufacturer*: The maker of the car (Chevrolet, Ford, Honda, Jeep)
5) *odometer*: How many miles the car has been driven

### Part 2a: Basic graph

Create the first graph for question 2 seen in Brightspace. Save it as **gg_q2a** and make sure to display it in the knitted document

```{r 2a}
used_cars <- read_csv('https://raw.githubusercontent.com/Shammalamala/DS-2870-Data-Sets/main/used%20cars.csv')
#create gg plot
gg_q2a <- ggplot(
  data = used_cars,
  mapping = aes(
    x = odometer,
    y = price,
    color = manufacturer#color based on manufacturer
  )
) +
  #add geom point and set color
  geom_point() +
  geom_smooth(
    data = used_cars,
    method = "lm", #linear
    formula = y~x,
    se = F,
    color = 'black', #black line and overrides manufacturer
    show.legend = F #gets rid of line on the legend
  )
  
gg_q2a
```

### Part 2b: Adding context

Using **gg_q2a**, add the title, subtitle, and caption and change/remove the labels for x, y, and color as seen in Brightspace. Change the legend to match Brightspace and move the legend to the top right corner of the plot.

Save it as **gg_q2b** display it in the knitted document.

```{r q2b}

gg_q2b <- gg_q2a +
  labs(
    x = "Mileage", #x axis title
    y = NULL, #get rid of y axis title
    title = 'Used Cars for Sale', #add title
    subtitle = 'Listed on Craigslist in 2023' #add subtitle
  ) +
  theme_classic() + #theme used for most features
  theme(
    plot.title = element_text(face = 'bold'), #bold plot and x axis title 
    axis.title.x = element_text(face = 'bold'),
    legend.title = element_blank(), #get rid of axis title
    legend.position = "inside", #move legend to inside the panel
    legend.position.inside = c(.9, .9) #adjust location
  )


gg_q2b
```

### Part 2c: Improving appearance

Make the final changes to the graph in **gg_q2b** that can be seen in Brightspace. Make sure to pay close attention to the color guide! 

The colors used are

Ford: #47a8e5
Chevrolet: #D1AD57
Honda: #CC0000
Jeep: #485F2B

```{r q2c}

gg_q2c <- gg_q2b + 
  scale_color_manual( values = c('#D1AD57', '#47a8e5', '#CC0000', '#485F2B'), #change legend colors
                      labels = c("Chevrolet", "Ford", "Honda", "Jeep")) + #capitalize legend names
  scale_y_continuous(labels = scales::label_dollar()) + #add dollar signs to y axis labels
  scale_x_continuous(labels = scales::label_comma()) #add commas to x axis labels


gg_q2c
```


## Question 3: Small Multiples

Create a set of 4 scatter plots with a fitted line in the same overall graph - 1 for each manufacturer. 

Each individual plot should have *odometer* on the x-axis, *price* on the y-axis, and *age* of the car represented by color.

```{r Q3}
used_cars$age <- (2023 - used_cars$year) #create age column
#create graph
gg_cars_grid <-  
  ggplot(
    data = used_cars, #data
    mapping = aes(
      x = odometer,
      y = price,
      color = age
    )
  ) + 
  labs(
    x = "Mileage", #x axis title
    y = NULL, #get rid of y axis title
    title = 'Used Cars for Sale', #add title
    subtitle = 'Listed on Craigslist in 2023' #add subtitle
  ) +
  theme_bw()+
  theme(
    plot.title = element_text(face = 'bold'), #bold title and x axis
    axis.title.x = element_text(face = 'bold')
  ) +
  geom_point() + #add scatter plot
  geom_smooth(
    method = "lm", #linear
    formula = y~x,
    se = F,
    color = 'black', #black line and overrides manufacturer
    show.legend = F #gets rid of line on the legend
  ) +
  facet_wrap( facets = vars( manufacturer ), nrow = 2,  scales = 'fixed' ) #facet based on manufacturer

gg_cars_grid
```

